---
title: "02_-_Exploring Data"
author: "Derek Sollberger"
format: 
  html:
    toc: true
---

## Load the Tidyverse

The `tidyverse` suite of packages are streamlined to make data science processes easier.  We can load the packages with the `library()` command.

```{r}
library("tidyverse")
```

## Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from Baseball Reference.  In particular, we are looking at the [team batting statistics](https://www.baseball-reference.com/leagues/majors/2021.shtml) for the 2021 season.

```{r}
df <- readr::read_csv("baseball_2021_batting.csv")
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
str(df, give.attr = FALSE)
```
Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
colnames(df)
```

## Measures of Centrality

Now let us practice computing means and medians of some of the numerical data.

```{r}
mean(df$HR)
```

During the 2021 season, MLB teams hit about 378 home runs on average.

```{r}
median(df$HR)
```

During the 2021 season, the median number of home runs per team was 198.

```{r}
mean(df$SB)
```

During the 2021 season, MLB teams stole about 141 bases on average.

```{r}
median(df$SB)
```

During the 2021 season, the median number of stolen bases per team was 75.
