---
title: "8: Rankings"
author: "Derek Sollberger"
date: "2023-11-09"
format: 
  html:
    toc: true
    theme: cerulean
---

# Start

## Load the Tidyverse

The `tidyverse` suite of packages are streamlined to make data science processes easier.  We can load the packages with the `library()` command.

The `gt` package allows researchers to quickly create tables that are easy to reorganize and are highly customizable.

```{r}
#| echo: true
#| message: false
#| warning: false
library("gt") #grammar of tables
library("tidyverse")
```

## Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.  For interest, this is a subset of the data from the 2004 to 2022 seasons.

```{r}
#| echo: true
#| eval: true
df <- readr::read_csv("baseball_2004_2022.csv") |>
  janitor::clean_names()
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
#| echo: true
#| eval: true
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
#| echo: true
#| eval: true
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
#| echo: true
#| eval: true
colnames(df)
```


# Examples

In each code block with multiple lines of code, we have "piped" together these lines of code with `|>` to connect the steps of the processes.  We can create a professional looking table by ending the chain with `gt()` ("grammar of tables").

## Highest Values

In the following example, here are the baseball teams (in the past 18 seasons) that have hit the most home runs in a season.

```{r}
#| echo: true
#| eval: true
df |>
  select(year_id, franch_id, hr) |> #select columns to keep
  slice_max(hr, n = 10) |> #finds 10 highest values by numerical variable
  gt() |> #makes nice tables
  tab_header(title = "Top 10 Teams by Home Runs")
```

From the `dplyr` part of the `tidyverse`, observe how the `slice_max` function will

* ask for a numerical variable
* arrange the data in descending order
* return the first $n$ rows

## Smallest Values

Similarly, the `slice_min` function will

* ask for a numerical variable
* arrange the data in ascending order
* return the first $n$ rows

In the following example, here are the baseball teams (in the past 18 seasons) that had the least earned run average in a season.

```{r}
#| echo: true
#| eval: true
df |>
  select(year_id, franch_id, era) |> #select columns to keep
  slice_min(era, n = 10) |> #finds 10 loweest values by numerical variable
  gt() |> #makes nice tables
  tab_header(title = "Top 10 Teams by ERA")
```

## Optional: Customization

`R` programmers like the `gt` package because we can customize the tables in many ways.  The documentation can be found here: [https://gt.rstudio.com/index.html](https://gt.rstudio.com/index.html)

Here is a slightly more extensive example.

```{r}
#| echo: true
#| eval: true
df |>
  select(year_id, franch_id, hr) |> #select columns to keep
  slice_max(hr, n = 10) |> #finds 10 highest values by numerical variable
  gt() |> #makes nice tables
  tab_header(title = md("Top 10 Teams by **Home Runs**")) |>
  tab_footnote(footnote = "2004 to 2022 seasons") |>
  tab_options(column_labels.background.color = "blue") |>
  cols_align(
    align = "center",
    columns = everything()
    ) |>
  tab_style(
    style = list(
      cell_fill(color = "yellow"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      columns = hr
    )
  )
```


# Exercises

Create a `gt` table that ranks one of the numerical variables from your sports data.  Your table has to include

* at least one numerical variable
* at least one categorical variable
* a title

```{r}

```

Create another `gt` table that ranks one of the numerical variables from your sports data.  Your table has to include

* at least one numerical variable
* at least one categorical variable
* a title

```{r}

```


# Wrap Up

* Click `Render` to create the HTML file
* Check the checkbox next to the HTML file (in the Files pane)
* Click the gear ("More"), and then click "Export"
* Upload the HTML file back into our CatCourses space