---
title: "12: Hypothesis Testing"
author: "Derek Sollberger"
date: "2022-11-15"
format: 
  html:
    toc: true
---

```{r}
library("infer")
library("moderndive")
library("patchwork") #easy graph layouts
library("tidyverse")
```

Source:  **Statistical Inference via Data Science:** *A Modern Dive into R and the Tidyverse*

* Chapter 9: Hypothesis Testing
* [https://moderndive.com/9-hypothesis-testing.html](https://moderndive.com/9-hypothesis-testing.html)

# Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.

```{r}

# df <- Teams |>
#   filter(yearID >= 1990) |>
#   filter(yearID < 2000)
# 
# readr::write_csv(df, "baseball_data_90s.csv")

df <- readr::read_csv("baseball_data_90s.csv")
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
colnames(df)
```

## Example 1

* explanatory variable (x): `lgID`
* response variable (y): `HR`

* null hypothesis: Both the National and American Leagues had the same average number of home runs.
* alternative hypothesis: The American League had a higher average number of home runs.

```{r}
# boxplot
df %>%
  filter(lgID == "NL" | lgID == "AL") %>%
  ggplot(aes(x = lgID, y = HR)) +
  geom_boxplot(aes(fill = lgID)) + 
  labs(title = "Home Runs by League",
       subtitle = "1990s baseball",
       caption = "Spark 01")
```

```{r}
# compute observed difference in means
obs_diff_means <- df %>%
  filter(lgID == "NL" | lgID == "AL") %>%
  specify(formula = HR ~ lgID) %>%
  calculate(stat = "diff in means", order = c("AL", "NL"))

# print
obs_diff_means
```


```{r}
# build a null distribution
null_distribution <- df %>%
  filter(lgID == "NL" | lgID == "AL") %>%
  specify(formula = HR ~ lgID) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("AL", "NL"))
```

```{r}
# visualize the p-value
visualize(null_distribution, bins = 10) + 
  shade_p_value(obs_stat = obs_diff_means, direction = "greater")
```

```{r}
# get p-value
null_distribution %>%
  get_p_value(obs_stat = obs_diff_means, direction = "greater")
```

Since the p-value is less than 0.05, we *reject* the claim that the two leagues had the same average number of home runs.

# Example 2

* explanatory variable (x): `franchID`
* response variable (y): `W`

* null hypothesis: Both the Dodgers and Giants averaged the same number of wins.
* alternative hypothesis: The Dodgers had a higher average number of wins than the Giants.

```{r}
# boxplot
df %>%
  filter(franchID == "LAD" | franchID == "SFG") %>%
  ggplot(aes(x = franchID, y = W)) +
  geom_boxplot(aes(fill = franchID)) + 
  labs(title = "Comparing the Dodgers and the Giants",
       subtitle = "1990s baseball",
       caption = "Spark 01")
```

```{r}
# compute observed difference in means
obs_diff_means <- df %>%
  filter(franchID == "LAD" | franchID == "SFG") %>%
  specify(formula = W ~ franchID) %>%
  calculate(stat = "diff in means", order = c("LAD", "SFG"))

# print
obs_diff_means
```

```{r}
# build a null distribution
null_distribution <- df %>%
  filter(franchID == "LAD" | franchID == "SFG") %>%
  specify(formula = W ~ franchID) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("LAD", "SFG"))
```

```{r}
# visualize the p-value
visualize(null_distribution, bins = 10) + 
  shade_p_value(obs_stat = obs_diff_means, direction = "greater")
```

```{r}
# get p-value
null_distribution %>%
  get_p_value(obs_stat = obs_diff_means, direction = "greater")
```
Since the p-value is greater than 0.05, we *fail to reject* the claim that the Dodgers and the Giants had the same average number of wins.
