---
title: "7: Linear Regression"
author: "Derek Sollberger"
date: "2023-09-14"
format: 
  html:
    toc: true
    theme: cerulean
---

## Load the Tidyverse

The `tidyverse` suite of packages are streamlined to make data science processes easier.  We can load the packages with the `library()` command.

```{r}
#| echo: true
#| message: false
#| warning: false
library("tidyverse")
```

# Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.

```{r}
#| echo: true
#| eval: true
df <- readr::read_csv("baseball_data_90s.csv") |>
  janitor::clean_names()
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
#| echo: true
#| eval: true
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
#| echo: true
#| eval: true
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
#| echo: true
#| eval: true
colnames(df)
```


# Linear Regression

A *scatterplot* plots points on a graph where both the horizontal and vertical axes are numerical variables.

```{r}
#| echo: true
#| eval: true
df |>
  ggplot(aes(x = r, y = w)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") + #makes the regression line
  labs(title = "MLB Data",
       subtitle = "Regression Line",
       caption = "seasons 1990 to 1999",
       x = "runs scored",
       y = "wins")
```

```{r}
#| echo: true
#| eval: true
# linear model (think "y = mx + b")
lin_fit <- lm(w ~ r, data = df)
m <- lin_fit$coefficients[2]
b <- lin_fit$coefficients[1]

df |>
  ggplot(aes(x = r, y = w)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") + #makes the regression line
  labs(title = "MLB Data",
       subtitle = paste0("y = ", round(m,4), "x + ", round(b, 4)),
       caption = "seasons 1990 to 1999",
       x = "runs scored",
       y = "wins")
```

# Making Predictions

Example: How many wins do we expect for a baseball team that scores 800 runs in a season?

```{r}
#| echo: true
#| eval: true
lin_fit <- lm(w ~ r, data = df)
predict(lin_fit, newdata = data.frame(r = 800))
```

We predict about 83.48 wins for a baseball team that scores 800 runs in a season.


# Another Example

Example: How many wins do we expect for a baseball team whose pitching ERA  (earned run average) is 4.00?

```{r}
#| echo: true
#| eval: true
# linear model (think "y = mx + b")
lin_fit <- lm(w ~ era, data = df)
m <- lin_fit$coefficients[2]
b <- lin_fit$coefficients[1]

df |>
  ggplot(aes(x = era, y = w)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue") + #makes the regression line
  labs(title = "MLB Data",
       subtitle = paste0("y = ", round(m,4), "x + ", round(b, 4)),
       caption = "seasons 1990 to 1999",
       x = "earned run average",
       y = "wins")
```

```{r}
#| echo: true
#| eval: true
lin_fit <- lm(w ~ era, data = df)
predict(lin_fit, newdata = data.frame(era = 4))
```

We predict about 80.76 wins for a baseball team whose season ERA is 4.00.


# Exercises

Load your data set (for your sport) using the `read_csv` command and save the data frame as `df`.

```{r}

```

Use the `head`, `str`, and `colnames` commands to look at your data set.

```{r}

```

```{r}

```

```{r}

```

Today, we are building linear regression models and making predictions in two scenarios from your sports data.

## Scenario 1

* describe what is on the $x$ axis: 
* describe what is on the $y$ axis: 

Build a scatterplot, and use a `geom_smooth` layer to draw a linear regression line.

```{r}

```

Use the `lm` and `predict` functions to carry out a prediction in this scenario (i.e. you choose a reasonable number for $x$).

```{r}

```

* describe your prediction: 

## Scenario 2

* describe what is on the $x$ axis: 
* describe what is on the $y$ axis: 

Build a scatterplot, and use a `geom_smooth` layer to draw a linear regression line.

```{r}

```

Use the `lm` and `predict` functions to carry out a prediction in this scenario (i.e. you choose a reasonable number for $x$).

```{r}

```

* describe your prediction: 


# Wrap Up

* Click `Render` to create the HTML file
* Check the checkbox next to the HTML file (in the Files pane)
* Click the gear ("More"), and then click "Export"
* Upload the HTML file back into our CatCourses space