---
title: "3: Summaries"
author: "Derek Sollberger"
date: "2023-08-31"
format:
  html:
    toc: true
    theme: cerulean
# format:
#   revealjs:
#     scrollable: true
---

Today we are going to try out convenient code packages to quickly explore our sports data sets.

# Example

To install a package, we use the `install.packages()` function in the console.  For example, when I want to install the `skimr` package, I use the following line of code.

```{r}
# install.packages("skimr")
```

```{r}
#| message: false
#| warning: false
# load the code packages
library("skimr")
library("tidyverse")
```

Since we have a CSV file, I recommend using the `read_csv()` function in the `readr` package (a part of the `tidyverse`).  At first, I like to save the data into a variable called `df` ("data frame").

```{r}
# load the data
df <- readr::read_csv("baseball_data_90s.csv")
```

Now, having read the `skimr` documentation, I can try out its summary function on the `df` data frame.

```{r}
skimr::skim(df)
```


# Exercise

Now it is your turn!  First, instead of `skimr`, *install* the package that has been assigned to your group.

* `DataExplorer::plot_intro()`
* `dataMaid::visualize()`
* `gtExtras::gt_plt_summary()`
* `SmartEDA::ExpData()`
* `summarytools::dfSummary()`

Next, load the code package using the `library()` function.

```{r}

```

Since we have a CSV file, I recommend using the `read_csv()` function in the `readr` package (a part of the `tidyverse`).  Use the file that is available for your group.

```{r}

```

Finally, look at the documentation for your group's code package, and use the summary function that has been recommended.

```{r}

```

# Wrap Up

* Click `Render` to create the HTML file
* Check the checkbox next to the HTML file (in the Files pane)
* Click the gear ("More"), and then click "Export"
* Upload the HTML file back into our CatCourses space