---
title: "13: Clustering"
author: "Derek Sollberger"
date: "2022-11-29"
format: 
  html:
    toc: true
---

```{r}
library("tidymodels")
library("tidyverse")
```

Source:  tidymodels

[https://www.tidymodels.org/learn/statistics/k-means/](https://www.tidymodels.org/learn/statistics/k-means/)

# Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.

```{r}

# df <- Teams |>
#   filter(yearID >= 1990) |>
#   filter(yearID < 2000)
# 
# readr::write_csv(df, "baseball_data_90s.csv")

df <- readr::read_csv("baseball_data_90s.csv")
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
colnames(df)
```

## Example of Clustering (old code)

```{r}
df_num <- df |>
  select(H, HR)
head(df_num)
```

```{r}
# scatterplot
df |>
  select(H, HR) |>
  ggplot(aes(x = H, y = HR)) +
  geom_point() +
  labs(title = "MLB Offense",
       subtitle = "1990s",
       caption = "Spark 01") +
  theme_minimal()
```

```{r}
# k-means clustering
clusters <- df |>
  select(H, HR) |>
  kmeans(centers = 5)
```

```{r}
# extraction
df_centers <- data.frame(x1 = clusters$centers[,1],
                         x2 = clusters$centers[,2])
df <- df |>
  mutate(cluster = clusters$cluster)
```

```{r}
df |>
  ggplot(aes(x = H, y = HR)) +
  geom_point() +
  
  # centers
  geom_point(aes(x = x1, y = x2),
             color = "red",
             data = df_centers,
             size = 5) +
  
labs(title = "MLB Offense",
       subtitle = "1990s",
       caption = "Spark 01") +
  theme_minimal()
```

```{r}
df |>
  ggplot(aes(x = H, y = HR)) +
  geom_point(aes(color = factor(cluster))) +
  
labs(title = "MLB Offense",
       subtitle = "1990s",
       caption = "Spark 01") +
  theme_minimal()
```


* **How many clusters** should we use?

## Clustering with TidyModels (modern code)

```{r}
# remember to use numerical columns
df_num <- df |>
  select(H, HR)

# try k = 1 cluster, k = 2 clusters, etc.
kclusts <- 
  tibble(k = 1:9) %>%
  mutate(
    kclust = map(k, ~kmeans(df_num, .x)),
    tidied = map(kclust, tidy),
    glanced = map(kclust, glance),
    augmented = map(kclust, augment, df_num)
  )
```

```{r}
# visualizing 9 possibilities
kclusts |>
  unnest(cols = c(augmented)) |>
  ggplot(aes(x = H, y = HR)) +
  geom_point(aes(color = .cluster), alpha = 0.75) +
  facet_wrap( ~ k) +
  theme_minimal()
```

```{r}
# Scree plot
kclusts |>
  unnest(cols = c(glanced)) |>
  ggplot(aes(k, tot.withinss)) +
  geom_line() +
  geom_point() +
  labs(title = "Scree Plot",
       subtitle = "for k-means clustering",
       caption = "Spark 01",
       x = "number of clusters",
       y = "total within SSE") +
  scale_x_continuous(breaks = 1:9) +
  theme_minimal()
```

After finding the "elbow" for the recommended number of clusters

```{r}
# visualizing clustering
kclusts |>
  unnest(cols = c(augmented)) |>
  filter(k == 3) |> #put number of clusters here
  ggplot(aes(x = H, y = HR)) +
  geom_point(aes(color = .cluster), alpha = 0.75) +
  labs(title = "MLB Offense",
       subtitle = "1990s",
       caption = "Spark 01") +
  theme_minimal()
```

# Clustering Workflow (old code)

(Just in case `tidymodels` doesn't install well on your computer.)

```{r}
# remember to use numerical columns
df_num <- df |>
  select(H, HR)

# allocate vector for tracking error amounts
tot_within_SSE <- rep(NA, 9)

# loop through k = 1 cluster, k = 2 clusters, etc.
for(k in 1:9){
  this_clustering <- kmeans(df_num, centers = k)
  tot_within_SSE[k] <- this_clustering$tot.withinss
}
```

```{r}
# make data frame (to prepare for ggplot)
df_scree <- data.frame(k = 1:9,
                       tot_within_SSE)

# scree plot
df_scree |>
  ggplot(aes(k, tot_within_SSE)) +
  geom_line() +
  geom_point() +
  labs(title = "Scree Plot",
       subtitle = "for k-means clustering",
       caption = "Spark 01",
       x = "number of clusters",
       y = "total within SSE") +
  scale_x_continuous(breaks = 1:9) +
  theme_minimal()
```

```{r}
# redo clustering one more time with recommended number of clusters
clusters <- kmeans(df_num, centers = 3)

# augment data frame
df_num <- df_num |>
  mutate(cluster = clusters$cluster)

# graph
df_num |>
  ggplot(aes(x = H, y = HR)) +
  geom_point(aes(color = factor(cluster)), alpha = 0.75) +
  labs(title = "MLB Offense",
       subtitle = "1990s",
       caption = "Spark 01") +
  theme_minimal()
```

