---
title: "11: Confidence Intervals"
author: "Derek Sollberger"
date: "2022-11-08"
format: 
  html:
    toc: true
---

```{r}
library("infer")
library("moderndive")
library("patchwork") #easy graph layouts
library("tidyverse")
```

Source:  **Statistical Inference via Data Science:** *A Modern Dive into R and the Tidyverse*

* Chapter 8: Bootstrapping and Confidence Intervals
* [https://moderndive.com/8-confidence-intervals.html](https://moderndive.com/8-confidence-intervals.html)

# Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.

```{r}

# df <- Teams |>
#   filter(yearID >= 1990) |>
#   filter(yearID < 2000)
# 
# readr::write_csv(df, "baseball_data_90s.csv")

df <- readr::read_csv("baseball_data_90s.csv")
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
colnames(df)
```

## One Sample

```{r}
HR_sample <- df |> select(HR)
```


```{r}
p1 <- HR_sample |>
  ggplot(aes(x = HR)) +
  geom_histogram(binwidth = 5, color = "black", fill = "blue") +
  labs(title = "Distribution of Home Runs",
       subtitle = "Based on a sample of data",
       caption = "Spark 01") +
  theme_minimal()

# print
p1
```

## Resampling

Using the available sample of data to fabricate another sample is called *resampling*.

```{r}
HR_resampled <- HR_sample |>
  sample_n(replace = TRUE, size = nrow(HR_sample))
```

```{r}
p2 <- HR_resampled |>
  ggplot(aes(x = HR)) +
  geom_histogram(binwidth = 5, color = "black", fill = "blue") +
  labs(title = "Distribution of Home Runs",
       subtitle = "Based on a sample of data",
       caption = "Spark 01") +
  theme_minimal()

# print
p2
```

```{r}
# patchwork
p1 + p2
```

## Resampling Many Times

```{r}
summary(HR_sample)
```


Suppose now that we have each person in a 23-student classroom repeat the act of drawing that home run data with replacement.

```{r}
HR_resampled_many <- HR_sample |>
  rep_sample_n(replace = TRUE, reps = 23, size = nrow(HR_sample))
```

```{r}
HR_resampled_many %>%
  group_by(replicate) %>%
  mutate(mean_HR = mean(HR)) %>%
  ungroup() %>%
  select(replicate, mean_HR) %>%
  distinct() %>%
  ggplot(aes(x = mean_HR)) +
  geom_histogram(binwidth = 1, color = "black", fill = "blue") +
  labs(title = "Resampling Results",
       subtitle = "N = 23 resamples",
       caption = "Source: Modern Dive")
```

Out of curiosity, let us push this process to $N = 1337$ resamples.

```{r}
HR_resampled_many <- HR_sample |>
  rep_sample_n(replace = TRUE, reps = 1337, size = nrow(HR_sample))
```

```{r}
HR_resampled_many %>%
  group_by(replicate) %>%
  mutate(mean_HR = mean(HR)) %>%
  ungroup() %>%
  select(replicate, mean_HR) %>%
  distinct() %>%
  ggplot(aes(x = mean_HR)) +
  geom_histogram(binwidth = 1, color = "black", fill = "blue") +
  labs(title = "Resampling Results",
       subtitle = "N = 23 resamples",
       caption = "Source: Modern Dive")
```

# Using the Infer package

The `infer` package streamlines the process of resampling (with replacement) to build a **bootstrap distribution**.

```{r}
bootstrap_distribution <- df |>
  specify(response = HR) |>
  generate(reps = 1337, type = "bootstrap") |>
  calculate(stat = "mean")
```

The folks who made the `infer` package provide a `visualize` helper function to quickly view the bootstrap distribution.

```{r}
visualize(bootstrap_distribution)
```

We can get the values of the endpoints of the confidence interval with the `get_ci` function.

```{r}
ci_endpoints <- bootstrap_distribution |> get_ci(level = 0.95, type = "percentile")

# print
ci_endpoints
```

We can also visualize the confidence interval on top of the bootstrap distribution.

```{r}
visualize(bootstrap_distribution) +
  shade_ci(ci_endpoints, color = "#DAA900", fill = "#002856") +
  labs(title = "Bootstrap Distribution of Home Runs",
       subtitle = "95% confidence interval shaded",
       caption = "Spark 01")
```

We are 95 percent confident that the true average number of home runs hit by a team in a season is in between 144 and 154 home runs.