---
title: "03_-_Boxplots"
author: "Derek Sollberger"
format: html
---

## Load the Tidyverse

The `tidyverse` suite of packages are streamlined to make data science processes easier.  We can load the packages with the `library()` command.

```{r}
library("tidyverse")
```

## Load the Data

Since we have CSV files (comma-separated values), the `read_csv()` command in the  `readr` package will be convenient here.  I tend to save data set into a variable `df` (stands for "data frame").

Today's data set comes from the `Lahman` package, which contains a lot of historical data about Major League Baseball.

```{r}
df <- readr::read_csv("baseball_teams_data.csv")
```

## Look at the Data

One way to get a quick look at the data set is with the `head()` command (outputs the first few rows and columns).

```{r}
head(df)
```

We can also look at the *structure* of a data frame with the `str()` command.  In particular, this view allows us to quickly see which variables are *numerical* and which are *categorical*.

```{r}
str(df, give.attr = FALSE)
```

Another convenient tool for our programming purposes is looking at the *column names* (i.e. variable names that we need to type later)

```{r}
colnames(df)
```

## A Starter Boxplot

```{r}
df |>
  ggplot(aes(y = W)) +
  geom_boxplot() +
  labs(title = "MLB Wins",
       subtitle = "2012 to 2021 seasons",
       caption = "Spark 01",
       y = "wins")
```

## Comparisons

```{r}
df |>
  ggplot(aes(x = lgID, y = R)) +
  geom_boxplot() +
  labs(title = "MLB Runs by League",
       subtitle = "2012 to 2021 seasons",
       caption = "Spark 01",
       x = "league",
       y = "runs scored")
```

## Grouping

```{r}
df |>
  ggplot(aes(x = franchID, y = W, group = franchID)) +
  geom_boxplot() +
  labs(title = "MLB Wins by Team",
       subtitle = "2012 to 2021 seasons",
       caption = "Spark 01",
       x = "team",
       y = "wins")
```

```{r}
df |>
  filter(franchID %in% c("ARI", "COL", "LAD", "SDP", "SFG")) |>
  ggplot(aes(x = franchID, y = W, group = franchID)) +
  geom_boxplot(aes(color = franchID, fill = franchID)) +
  
  # https://teamcolorcodes.com/mlb-color-codes/
  scale_color_manual(values = c("#E3D4AD", "#C4CED4", "#A5ACAF", "#FFC425", "#27251F")) +
  scale_fill_manual(values = c("#A71930", "#333366", "#005A9C", "#2F241D", "#FD5A1E")) +

  labs(title = "MLB Wins by Team",
       subtitle = "2012 to 2021 seasons",
       caption = "Spark 01",
       x = "team",
       y = "wins") +
  theme_minimal()
```